---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# redlistr

[![Build Status](https://travis-ci.org/red-list-ecosystem/redlistr.svg?branch=master)](https://travis-ci.org/red-list-ecosystem/redlistr)

## Overview
The `redlistr` package was developed to assist users conduct assessments for the
IUCN Red List of Ecosystems in `R`. It is also useful for users interested in
conducting assessments for the Red List of Threatened Species. Assessments of
ecosystems under the IUCN Red List of Ecosystems criteria require calculation of
standardised metrics that were developed to objectively assess risk to ecosystem
([Keith et al. 2013](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0062111)).
This package was designed to assist in the calculation of these metrics, 
including two methods of calculating the rate of distirbutional decline: 
Absolute Rate of Decline (ARD) and Proportional Rate of Decline (PRC). An 
additional metric: the Annual Rate of Change (ARC), which uses a compound 
interest law to determine the instantaenous rate of change ([Puyravaud 2003](http://www.sciencedirect.com/science/article/pii/S0378112702003353)) is
also included.

Also included are the two standard measures of the size of an 
ecosystems' geographic distribution specified in the red list of ecosystems
guidelines ([Bland et al. 2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf)).
These are the Extent of Occurrence (EOO) and Area of Occupancy (AOO). As many of
these measures are also useful for assessing species under the IUCN Red List of
Threatened Species criteria, we expect this package will also be useful for
assessors conducting species assessments.

In conducting an assessment with this package, we assume that you are familiar 
with IUCN red listing protocols. In particular, you should consult the IUCN 
guidelines for both of the red lists, which are the definitive sources of all 
information required to ensure consistent application of IUCN criteria (IUCN, 
[2015](link here); [Bland et al. 2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf)).
In addition, the papers by Keith et al.
([2013](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0062111))
and Rodriguez et al.
([2015](https://www.iucn.org/sites/dev/files/content/documents/rodriguez-etal-2015-a-practical-guide-iucn-rle_erratum.pdf))
are particularly useful for navigating the IUCN Red List of Ecosystems criteria.
A range of important resources, including software tools and guiding material is
available on the IUCN Red List of Ecosystems 
[website](https://www.iucnrle.org/). There is also plenty of handy material for 
assessing species available on the IUCN Red List of Threatened Species 
[website](http://www.iucnredlist.org).

We also assume that you are reasonably familiar with the `R` programming
language, and have some experience in conducting analyses of vector and raster
data within the `R` environment. Of particular use will be familiarity with the
`raster`, `sp` and `rgeos` packages. This is certainly not a prerequisite, but
this package is built upon many of the functions available in these three
packages.

For a more detailed tutorial explaining how to use this package, please refer to
the 'Example red list of ecosystems assessment' vignette available with the
package.

Lastly, this is a work in progress and we aim to continually add new functions to 
newer versions of package. Suggestions are welcomed, as would offers for 
collaborative development.

## Installation
Currently, `redlistr` can only be installed via GitHub, which requires the
`devtools` package:
```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("red-list-ecosystem/redlistr")
```

