---
title: "Get started"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

## What is this package for?

`redlistr` is a package to assess spatial criteria of the [IUCN Red List of
Ecosystems](https://iucnrle.org/) (RLE) and [IUCN Red List of
Threatened Species](https://www.iucnredlist.org/) (RLTS). 

The package has functions for calculating three metrics that are useful for different subcriteria of RLE and RLTS:

<p style="text-align:center;">

| Metrics  | main `redlistr` function | RLE subcriteria | RLTS subcriteria |
| ------------- | -------------  | -------------  | ------------- | 
| Declines in distribution   | `getDeclineStats` | A1, A2, A3|  A1c, A2c, A3c, A4c |
| Extent of Occurrence (EOO) | `getEOO` | B1 | B1  |
| Area of Occupancy (AOO) | `getAOO` | B2 | B2, D2 |
</p>

## Input data formats

Functions in this package expect spatial data as input. They should represent a valid estimate of species or ecosystem distribution at one or multiple points in time.

Since version 3.0.0 the `redlistr` functions will accept different formats for the input vector and raster data. The spatial data need to be read in R using functions from external packages, either `sf` or `terra`:
 
<p style="text-align:center;">

| Data type | Example formats | R package | R class | Example functions |
| ------------- | -------------  | -------------  | ------------- | 
| Vector   | Shapefile, Geopackage, GeoJSON, GDB | `sf` | `sf` |  `sf::read_sf` |
| Vector   | Shapefile, Geopackage, GeoJSON, GDB | `terra` | `SpatVect` |  `terra::vect` |
| Raster   | GeoTIFF | [`terra`](https://rspatial.github.io/terra/) | `SpatRast` |  `terra::rast` |
</p>

## Outputs of the functions 

The output of the `redlistr` functions will be R objects with class attributes. There are `summary` and `print` functions for these classes.

## Using `redlistr`

A simple workflow for using `redlistr` consist of importing and preparing the input data, running the main function and using the `summary` function to get the main information from the resulting object.

```{r Loading packages, message=FALSE, echo=FALSE, results="hide"}
library(terra)
library(redlistr)
raster_file <- system.file("extdata", "example_distribution_2000.tif", 
                                    package = "redlistr")
```

For example, if the path of the spatial data is in a R character variable names `raster_file` and the data already has a valid Coordinate Reference System in meters. These three lines of code will create the EOO polygon and print the results of the area in square kilometers.

```{r Loading our example distributions}
ecosystem_map <- rast(raster_file)
ecosystem_EOO <- getEOO(ecosystem_map)
summary(ecosystem_EOO)
```

