---
title: "Calculating spatial metrics for IUCN red list assessments"
author: "Calvin K.F. Lee, Nicholas J. Murray"
date: "2017-07-07"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating spatial metrics for IUCN red list assessments}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

## 1. Introduction

In this vignette, we provide an example of using the tools within
`redlistr` to assess the spatial criteria of the [IUCN Red List of
Ecosystems](https://iucnrle.org/). These criteria assess the change in
the extent of an ecosystem over time (Criterion A) and properties of the
geographic distribution size (Criterion B). Both of these criteria
require the use of maps of ecosystem distributions.

We begin by loading the packages we need for the assessment. Please
ensure that these are already installed.

```{r Loading packages, message=FALSE}
library(sf)
library(terra)
library(redlistr)
```

## 2. Importing data

We first import the spatial data we want to analyse. In this vignette,
we will be using two example distributions included with the package.
They are mangrove distributions from the northern regions of Western
Port Bay, Victoria, Australia. The first distribution is from Giri et
al.,
([2011](https://onlinelibrary.wiley.com/doi/10.1111/j.1466-8238.2010.00584.x/abstract)),
and represents mangrove distribution in 2000. The second distribution is
a classification map generated using the `XGBoost` package and data from
Landsat 8, and represents mangrove distribution in 2017. Both of these
geographic distributions are stored as rasters, and a raster value of 1
denotes presence.

```{r Loading our example distributions}
mangrove.2000 <- rast(system.file("extdata", "example_distribution_2000.tif", 
                                    package = "redlistr"))
mangrove.2017 <- rast(system.file("extdata", "example_distribution_2017.tif", 
                                    package = "redlistr"))
```

### Importing data as rasters

To import your own raster data, you can use the `terra::rast` function,
which can handle many different types of georeferenced spatial data.
More information on the `terra` package can be found
[here](https://rspatial.github.io/terra/).

### Importing data as shapefile or .KML file

To import data as a shapefile (.shp), we can use the `sf::st_read`
function.

```{r Importing shapefile, eval=FALSE}
library(sf)
my.shapefile <- st_read('./path/to/folder/', 'shapefile.shp')
my.KML.file <- st_read('./path/to/folder/kmlfile.kml')
```

Once these files are imported, they can then be converted into raster
format via the `terra::rast` or `terra::rasterize` functions if desired.
However, many of the functions in the package will accept shape files
directly.

### Plotting out data

We plot the data to view the two mangrove distributions to get a general
idea of how the distribution might have changed over time.

```{r Plotting the two rasters, fig.show='hold', fig.width=7, fig.height=7}
plot(mangrove.2000, col = "grey30", legend = FALSE, main = "Mangrove Distribution")
plot(mangrove.2017, add = T, col = "darkorange", legend = FALSE)
```

We can see that there has been some change in mangrove cover, although
the losses appear to be relatively minor.

At this stage it is important to check that the data are georectified
(in the right location on Earth). This can be achieved by simply
plotting the map and checking the coordinates. Otherwise, check the
distribution maps against satellite images, perhaps by using packages
such as `ggmap`, `plotgooglemaps`, `googleVis` etc. It is also important
to check that the data are suitable for the task at hand.

## 3. Basic information for the data

We now use some of features of `redlistr` to get some basic information
on the ecosystem distribution. However, before proceeding, we check
their resolution (grain size), extent, and spatial properties.
Specifically, the coordinate reference system (CRS) of the datasets must
be measured in metres, and must be consistent between layers you wish to
compare.

```{r Checking CRS}
is.lonlat(mangrove.2000) 
is.lonlat(mangrove.2017)
# If TRUE, they must be transformed to a projected coordinate system

crs(mangrove.2000) == crs(mangrove.2017)
```

Reprojection can be done by using the `terra::project` function. Be sure
to choose a coordinate reference system that is appropriate to your
study system.

We start by calculating the area of the ecosystem. Area is calculated
here using the pixel count method, where the number of pixels classified
as the ecosystem of interest is multiplied by the area of each pixel,
giving the area of the ecosystem distribution. The area calculated is
provided in square kilometres.

```{r Calculate area of rasters}
a.2000 <- getArea(mangrove.2000)
a.2000
a.2017 <- getArea(mangrove.2017)
a.2017

# you can also calculate areas for both layers at once by combining them into a raster stack:

a <- getArea(c(mangrove.2000, mangrove.2017))
```

The results: at time 1 (2000), the ecosystem was `r a.2000$area` km2,
and at time 2, (2017) the ecosystem was `r a.2017$area` km2.

### Rasters with multiple ecosystems or layers

The `getArea` function is capable of processing rasters with multiple
ecosystems and layers. Layers usually represent time slices. The
function will return a table of areas, with a row for each ecosystem
represented in the raster, and a separate row for each layer if there
are multiple layers.

If you prefer to focus on a single ecosystem or raster value, and you do
not know the value which represents your target ecosystem, you can plot
the ecosystem and use the `terra::click` function. Once you know the
value which represents your target ecosystem, you can create a new
raster object including only that value, or for smaller rasters, simply
filter your output table to the value(s) of interest. Here's an example
with a dummy raster.

```{r rasters with multiple values, eval=FALSE}
# Load dummy multilayer raster for example
rstack <- rast(system.file("extdata", "example_raster_stack.tif", 
                                    package = "redlistr"))

# If the target ecosystem is represented by value = 5
r.bin <- rstack == 5 # Has values of 1 and 0
# Convert 0s to NAs, if desired. 
values(r.bin)[values(r.bin) != 1] <- NA
# get area
getArea(r.bin)

# Alternatively: 
getArea(rstack) |> 
  dplyr::filter(value == 5)
 # note this alternative is slower for large rasters. 
```

### Polygon data with multiple ecosystems

If your input data is in polygon format, you can enter the name of the
column containing ecosystem labels into the names_from argument. If left
empty, for polygon inputs the function will treat all polygons as part
of a single ecosystem type.

```{r polygon data for multiple ecosystems, eval = F}
# need example polygon data
#TODO

```

## 4. Assessing Criterion A

The IUCN Red List of Ecosystems criterion A requires estimates of the
magnitude of change over time. This is typically calculated over a 50
year time period (past, present or future) or against a historical
baseline ([Bland et al.,
2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf)).
The first step towards achieving this change estimate over a fixed time
frame is to assess the amount of change observed in your data.

### Area change

Area change can be calculated for one of two scenarios: multiple
ecosystems over two time slices using `getAreaChange()`, or one or more
ecosystems over multiple time slices using `getAreaTrend()`.

The `getAreaChange` function takes two inputs in the same format. It
calculates the difference in area between two datasets, which can
contain multiple ecosystems. It can accept data from `data.frame` such
as those output by the `getArea()` function, `SpatRaster`, `SpatVector`
or `sf` objects with POLYGON geometries. Make sure to provide area
measure in km2 if you are using a data.frame as the input. Note that
when providing data frame and polygon inputs, the column name with
unique ecosystem labels must be included in the function call. If the
column name is different in the two inputs, both names must be provided
in the same order as the input datasets.

```{r Using getAreaChange}
# input rasters
area.change <- getAreaChange(mangrove.2000, mangrove.2017)

# input data frames 
# value is the column that identifies the ecosystem in both tables.

area.change <- getAreaChange(a.2000, a.2017, value) 

# input polygons 
# TODO

```

The `getAreaTrend` function takes one multilayer raster or a list of sf
or SpatVector objects. It calculates the area for each layer or list
element, for each ecosystem. The function returns a list of `trend`
class objects, with one element for each ecosystem or raster value in
the input data.

```{r}
#input multilayer raster 
# returns a list of trends, one per ecosystems
area.trend <- getAreaTrend(rstack)

# summarise the change for ecosystem 1 (includes plot of area over time)
summary(area.trend[[1]])

# plot the diff raster for ecosystem 1
plot(area.trend[[1]])

#input polygons
#TODO
```

### Rate of change

In the Red List of Ecosystems, two methods are suggested to determine
the rate of decline of an ecosystem, each of which assumes a different
functional form of the decline ([Bland et al.,
2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf)).
The proportional rate of decline (PRD) is a fraction of the previous
year's remaining area, while the absolute rate of decline (ARD) is a
constant fraction of the area of the ecosystem at the beginning of the
decline ([Bland et al.,
2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf)).
These rates of decline allow the use of two or more data points to
extrapolate to the full 50 year timeframe required in an assessment.

The annual rate of change (ARC) uses a compound interest law to
determine the instantaneous rate of change ([Puyravaud
2003](https://www.sciencedirect.com/science/article/pii/S0378112702003353)).

To estimate the rate of changing using these methods, we use the
`getDeclineStats` function.

```{r Using getDeclineStats}
decline.stats <- getDeclineStats(a.2000$area, a.2017$area, 2000, 2017, 
                                 methods = c('ARD', 'PRD', 'ARC'))
decline.stats
```

Each method represent a different shape of decline. For further
information about the choice of each of these methods to extrapolate
refer to the IUCN Red List of Ecosystems guidelines ([Bland et al.,
2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf)).

Now, it is possible to extrapolate, using only two estimates of an
ecosystems' area, to the full 50 year period required for a Red List of
Ecosystems assessment.

```{r Estimating future area}
extrapolated.area <- futureAreaEstimate(a.2000$area, year.t1 = 2000, 
                                        ARD = decline.stats$ARD, 
                                        PRD = decline.stats$PRD, 
                                        ARC = decline.stats$ARC, 
                                        nYears = 50)
extrapolated.area
```

50 years from our first estimate is the year
`r extrapolated.area$forecast.year`.

As we included all three methods of calculating rate of decline
currently included in the package, the results produced shows three
estimated areas under the various decline scenarios. It is important to
note that this relatively simple exercise is founded on assumptions that
should be fully understood before submitting your ecosystem assessment
to the IUCN Red List of Ecosystems Committee for Scientific Standards.
Furthermore, the guidelines suggest using area estimates from more than
two time points to estimate change, and providing a measure of
uncertainty if possible. Please see the guidelines ([Bland et al.,
2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf))
for more information.

If we were to use the Proportional Rate of Decline (PRD) for our example
assessment, our results here will be suitable for criterion A2b (Any 50
year period), and the percent loss of area is:

```{r Percent loss}
predicted.percent.loss <- (extrapolated.area$A.PRD.t3 - a.2000$area)/a.2000$area * 100
predicted.percent.loss
```

## 4. Assessing Criterion B (distribution size)

Criterion B utilizes measures of the geographic distribution of an
ecosystem type to identify ecosystems that are at risk from catastrophic
disturbances. This is done using two standardized metrics: the extent of
occurrence (EOO) and the area of occupancy (AOO) [(Gaston and Fuller,
(2009)](https://onlinelibrary.wiley.com/doi/10.1111/j.1365-2664.2008.01596.x/full),
[Keith et al.,
(2013)](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0062111)).
It must be emphasised that EOO and AOO are not used to estimate the
mapped area of an ecosystem like the methods we used in Criterion A;
they are simply spatial metrics that allow us to standardise an estimate
of risk due to spatially explicit catastrophes [(Murray et al.,
(2017)](https://onlinelibrary.wiley.com/doi/10.1111/ddi.12533/abstract).
Thus, it is critical that these measures are used consistently across
all assessments, and the use of non-standard measures invalidates
comparison against the thresholds. Please refer to the guidelines
([Bland et al.,
2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf))
for more information on AOO and EOO.

### Subcriterion B1 (calculating EOO)

For subcriterion B1, we will need to calculate the extent of occurrence
(EOO) of our data. We begin by creating the minimum convex polygon
enclosing all occurrences of our focal ecosystem. There are some handy
summary functions for the EOO object that is returned.

```{r Make and view EOO, fig.width=7, fig.height=7}
EOO.polygon <- getEOO(mangrove.2017)
summary(EOO.polygon)
plot(EOO.polygon)
```

We can then extract the area of the EOO polygon, if needed.

```{r Calculating EOO area}
EOO.area <- getAreaEOO(EOO.polygon)

#or, alternately, extract the EOO directly from the object. 
EOO.area <- EOO.polygon@EOO
```

The calculated EOO area for this subset of mangroves in Victoria is
`r EOO.area` km2. This result can then be combined with additional
information required under B1(a-c) to assess the ecosystem under
subcriteria B1.

### Subcriterion B2 (calculating AOO)

For subcriterion B2, we will need to calculate the number of 10x10 km
grid cells occupied by our distribution. The `getAOO()` function uses a
random grid search to find the optimal grid with the fewest cells. For
ecosystems that are contained in more than 60 grid cells, the default
grid is used, as a random grid search is unlikely to yield results near
the IUCN Criteria thresholds.

```{r Creating AOO grid, fig.width=7, fig.height=7}
AOO.grid <- getAOO(mangrove.2017, grid.size = 10000)

# summary
summary(Aoo.grid)

# handy plotting function
plot(AOO.grid)

```

We can run some diagnostics to ensure we have the best grid. The
jitter-plot (jplot) shows how the number of cells decreased as the
random search iterations increased. The plot should fall to the minimum
number early and stay flat. You can also plot a histogram of the AOO
values generated with the random grid search iterations, and this should
look like a bell curve. If these plots look unconvincing, you can
increase the number of iterations with the argument `n_jitter` directly
in the `getAOO()` function.

```{r Getting number of AOO grids}

# jplot
jplot(AOO.grid)

# histogram of values generated with each rep
hist(AOO.grid@AOOvals)

# increase the number of iterations, if desired
AOO.grid <- getAOO(mangrove.2017, grid.size = 10000, n_jitter = 150)
```

#### One percent rule

As per the IUCN guidelines, the cells containing the smallest areas of
the ecosystem, and collectively adding to less than 1% of the
ecosystem's total area, are not counted by default. This prevents tiny,
far-flung fragments of an ecosystem from inflating the AOO. This option
can be turned off using the argument, `bottom.1pct.rule = FALSE`, or the
percentage changed using the `percent` argument.

Here, we demonstrate the differences between including, or not including
the one percent rule:

```{r One percent grid, fig.width=7, fig.height=7}
AOO.grid.one.percent <- getAOO(mangrove.2017, grid.size = 10000, 
                                    bottom.1pct.rule = F, n_jitter = 150)

par(mfrow = c(2,1))
plot(AOO.grid.one.percent)
plot(AOO.grid)
par(mfrow = c(1,1))


```

## 5. Final words

We have demonstrated in this vignette a typical workflow for assessing
an example ecosystem under the RLE. The methods outlined here can easily
be adapted for the Red List of Threatened Species as well, with slight
adjustments in the parameters (specifically the grid.size parameter for
`getAOO` or `gridUncertainty`). We hope that with this package, we help
ensure a consistent implementation for both red lists criteria,
minimising any misclassification as a result of using different software
and methods.
