---
title: "Calculating spatial metrics for red list of ecosystems assessment"
author: "Calvin K.F. Lee, Nicholas J. Murray"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating spatial metrics for red list of ecosystems assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## 1. Introduction
In this vignette, we will be providing an example of using the tools within
`redlistr` aimed at assessing the spatial criteria of the IUCN Red List of
Ecosystems. These criteria assess the change in the extent of an ecosystem over
time (Criterion A) and properties of the geographic distribution size (Criterion
B). Both of these criteria require the use of maps of ecosystem distributions.

We begin by loading the packages we will need for the assessment. Please ensure
that these are already installed.
```{r Loading packages, message=FALSE}
library(rgeos)
library(redlistr)
```

## 2. Importing data
We will first have to import the data we want to analyse. In this vignette, we 
will be using two example distributions included with the package. They are 
mangrove distributions from the northern regions of Western Port Bay, Victoria,
Australia. The first distribution is from Giri et al., 
([2011](http://onlinelibrary.wiley.com/doi/10.1111/j.1466-8238.2010.00584.x/abstract)),
and represents mangrove distribution in 2000. The second distribution is a 
classification map generated using the `XGBoost` package and data from Landsat 
8, and represents mangrove distribution in 2017. Both of these distributions are
stored as rasters, and a raster value of 1 denotes presence.

To import your own data, you can use the `raster` function in the `raster`
package, which can handle many different types of georeferenced spatial data.
More information on the `raster` package can be found
[here](https://cran.r-project.org/web/packages/raster/raster.pdf).

```{r Loading our example distributions}
mangrove.2000 <- raster(system.file("extdata", "example_distribution_2000.tif", 
                                    package = "redlistr"))
mangrove.2017 <- raster(system.file("extdata", "example_distribution_2017.tif", 
                                    package = "redlistr"))

```

We can plot the data to view the two distributions to get a general idea of how
the distribution might have changed over time.

```{r Plotting the two rasters, fig.show='hold', fig.width=7, fig.height=7}
plot(mangrove.2000, col = "grey30", legend = FALSE, main = "Mangrove Distribution")
plot(mangrove.2017, add = T, col = "darkorange", legend = FALSE)
```

We can see that there has been some decrease in mangrove cover, though the loss
does not seem to be too extensive.

At this stage it is also important to check that data are at least in the right
location. This can be acheived by simply plotting the map and checking the
coordinates, or check the distribution maps against satellite images. This can
be acheived using packages such as `ggmap`, `plotgooglemaps`, `googleVis` etc.
It is also important to check that the data are suitable for the task at hand.

## 3. Basic information for the data
We can now use some of features of `redlistr` to get some basic information on
the ecosystem. However, before proceeing it is wise to check that the data and
their respective resolutions (grain size), and spatial properties. Specifically,
the coordinate reference system (CRS) of the datasets must be consistent as we
are comparing their areas.

```{r Checking CRS and resolutions}
crs(mangrove.2000)@projargs == crs(mangrove.2017)@projargs
```

It should be noted here that for `redlistr` to function properly, the unit of
measurement in the projection used should be metres.

We start by calculating the area of the dataset. Area is calculated here using 
the pixel count method, where the number of pixels classified as the ecosystem 
of interest is multiplied by the area of each pixel is used to determine the
area of the whole ecosystem. The area calculated is provided in square
kilometres.

```{r Calculate area of rasters}
a.2000 <- getArea(mangrove.2000)
a.2000
a.2017 <- getArea(mangrove.2017)
a.2017
```

The results: at time 1 (2000), the ecosystem was `r a.2000` km^2^, and at time
2, (2017) the ecosystem was `r a.2017` km^2^.

## 4. Assessing Criterion A

The IUCN Red List of Ecosystems criterion A requires estimates of the magnitude of
change over time. This is typically calculated over a 50 year time period (past,
present or future) or against a historical baseline ([Bland et al., 2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf)).
The first step towards acheiving this change estimate over a fixed time frame is
to assess the amount of change observed in your data.

### Area change

Area change can be calculated using the `getAreaLoss` function. It simply
calculates the difference in area between two datasets. It can accept data from
numbers, `RasterLayers` or `SpatialPolygons`. Make sure to use area measure in
km^2^ if you are using numbers as the input.

```{r Using getAreaChange}
area.lost <- getAreaLoss(a.2000, a.2017)
# getAreaLoss(mangrove.2000, mangrove.2017) generates identical results

```

### Rate of change

In the Red List of Ecosystems, two methods are suggested to determine the rate
of decline of an ecosystem, each of which assumes a different functional form of
the decline (IUCN, 2015).  The proportional rate of decline (PRD) is a fraction
of the previous year's remaining area, while the absolute rate of decline (ARD)
is a constant fraction of the area of the ecosystem at the beginning of the
decline (IUCN, 2015). These rates of decline allow the use of two or more data 
points to extrapolate to the full 50 year timeframe required in an assessment.

The annual rate of change (ARC) uses a compound interest law to determine the 
instantaneous rate of change ([Puyravaud 2003](http://www.sciencedirect.com/science/article/pii/S0378112702003353)). 

To estimate the rate of changing using these methods, we use the
`getDeclineStats` function. 

```{r Using getDeclineStats}
decline.stats <- getDeclineStats(a.2000, a.2017, 2000, 2017, 
                                 methods = c('ARD', 'PRD', 'ARC'))
decline.stats
```

Each method represent a different shape of decline. For further information about
the choice of each of these methods to extrapolate refer to the IUCN Red List of
Ecosystems guidelines ([Bland et al., 2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf)).

Now, it is possible to extrapolate, using only two estimates of an ecosystems'
area, to the full 50 year period required for a Red List of Ecosystems
assessment.

```{r}
extrapolated.area <- futureAreaEstimate(a.2000, year.t1 = 2000, 
                                        ARD = decline.stats$ARD, 
                                        PRD = decline.stats$PRD, 
                                        ARC = decline.stats$ARC, 
                                        nYears = 50)
extrapolated.area
```

This tells us that 50 years from our first estimate in the year 
`r extrapolated.area$forecast.year`.

As we included all three methods of calculating rate of decline currently included in
the package, the results produced shows three estimated areas under the various
decline scenarios. It is important to note that this relatively simple exercise is founded on 
assumptions that should be fully understood before submitting your ecosystem 
assessment to the IUCN Red List of Ecosystems Committee for Scientific
Standards. Please see the guidelines ([Bland et al., 2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf))
for more information.

## 4. Assessing Criterion B (distribution size)

Criterion B utilizes measures of the geographic distribution of an ecosystem 
type to identify ecosystems that are at risk from catastrophic disturbances. 
This is done using two standardized metrics: the extent of occurrence (EOO) and 
the area of occupancy (AOO) [(Gaston and Fuller, (2009)](http://onlinelibrary.wiley.com/doi/10.1111/j.1365-2664.2008.01596.x/full), 
[Keith et al., (2013)](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0062111)). 
It must be emphasised that EOO and AOO are not used to estimate the mapped area
of an ecosystem like the methods we used in Criterion A; they are simply spatial
metrics that allow us to standardise an estimate of risk spreading. Thus, it is 
critical that these measures are used consistently across all assessments, and
the use of non-standard measures invalidates comparison against the thresholds. 
Please refer to the guidelines ([Bland et al., 2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf))
for more information on AOO and EOO.

### Subcriterion B1 (calculating EOO)
For subcriterion B1, we will need to calculate the extent of occurrence (EOO) of our
data. We begin by creating the minimum convex polygon enclosing all occurrences
of our data.
```{r Make EOO, fig.width=7, fig.height=7}
EOO.polygon <- makeEOO(mangrove.2017)
plot(EOO.polygon)
plot(mangrove.2017, add = T, col = "green", legend = FALSE)
```

We can then calculate the area of the EOO polygon.

```{r Calculating EOO area} 
EOO.area <- getAreaEOO(EOO.polygon)
EOO.area
```

The calculated EOO area for this subset of mangroves in Victoria is `r EOO.area`
km^2^. This result can then be combined with additional information required
under B1(a-c) to assess the ecosystem under subcriteria B1.

### Subcriterion B2 (calculating AOO)
For subcriterion B2, we will need to calculate the number of 10x10 km grid cells
occupied by our distribution. We begin by creating the appopriate grid cells.
```{r Creating AOO grid, fig.width=7, fig.height=7}
AOO.grid <- makeAOOGrid(mangrove.2017, grid.size = 10000, 
                        min.percent.rule = T, percent = 1)

plot(AOO.grid)
plot(mangrove.2017, add = T, col = "green", legend = FALSE)
```

In addition to the size of the grids used (which will be different for species
assessments), there is also an option in the `makeAOOGrid` function to specify
whether a minimum of percent of the grid cell area must be occupied before they
are included as an AOO grid. Typically, this is used for ecosystems with highly
skewed distribution of patch sizes ([Bland et al., 2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf)).
You can also modify the actual percentage required before an AOO is included in
the `percent` parameter.

Finally, we can use the created grid to calculate the AOO for our mangroves.
```{r Getting number of AOO grids}
n.AOO <- length(AOO.grid)
n.AOO

# the getAOO function can also be used to directly get the AOO
# n.AOO <- getAOO(mangrove.2017, grid.size = 10000, 
#                 min.percent.rule = T, percent = 1)
```


#### Grid uncertainty functions
Although we have a number for AOO `r n.AOO`, this may not be the actual minimum
AOO for this ecosystem, as the placement of the AOO grids will influence this
number. To take this into account, we have included a few additional functions.
We will be using `gridUncertainty` in this vignette.

`gridUncertainty` works by moving the AOO grid created systematically (with a
small random movement around fixed points), and continues looking for a minimum
AOO until additional shifts no longer produce improved results.

```{r gridUncertainty}
gU.results <- gridUncertainty(mangrove.2017, 10000,
                              n.AOO.improvement = 5, 
                              min.percent.rule = T, percent = 1)



```






