---
title: "Calculating spatial metrics for IUCN RLTS assessments"
author: "José R. Ferrer-Paris"
date: "2026-03-13"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating spatial metrics for IUCN RLTS assessments}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

## Introduction

In this vignette, we provide an example of using the tools within
`redlistr` to assess two criteria of the [IUCN Red List of
Threatened Species](https://www.iucnredlist.org). 

We will use a real life example and work through four steps that are necessary to evaluate these criteria using R.

In this vignette we will start with some context about our target species, then we will import a selection of R packages that will be helpful for different tasks, and then we proceed with the four steps of: data import, data preparation, calculation of metrics and application of thresholds. 

## The purple copper butterfly in Australia

_Paralucia spinifera_ Edwards and Common, 1978 is a small butterfly in the Lycaenidae Family. It has a purple and green iridescent upper surface and muted brown, black and grey undersides.

This species is known only from a small number of sites east of Bathurst. Its distribution is believed to be restricted by abiotic factos (locations above 900m elevation) and biotic interactions with a specific host plant (_Bursaria spinosa_) and attendant ant (_Anonychomyrma itinerans_.

Adults fly from late September to late November. Eggs are laid singly or in small groups. Eggs hatch in 14-17 days and the larval period lasts 6-8 weeks. Pupation takes place between December to late February in ants nests at the base of the host plant; pupae remain in the ants nest until the following spring.

The survival of Paralucia spinifera is threatened by habitat disturbance like grazing of the host plant by sheep, goats and cattle, clearing for the establishment of pine plantations, invasion of exotic weeds and activities of feral pigs. 

The latest conservation assessment was conducted in 2024:

Conservation Assessment of the purple copper butterfly Paralucia spinifera Edwards and Common, 1978 (Lycaenidae) Matt Saunders 15/08/2024 Conservation Policy and Programs Division NSW Department of Climate Change, Energy, the Environment and Water. [link](https://www.environment.nsw.gov.au/topics/animals-and-plants/threatened-species/nsw-threatened-species-scientific-committee/nsw-threatened-species-scientific-committee-publications/assessment-reports)

Paralucia spinifera was found to be Endangered under IUCN Criterion B1b(ii,iii,iv,v)c(iv)+2b(ii,iii,iv,v)c(iv). 

## Loading packages in R

We begin by loading the packages we need for data import and for calculation of the spatial metrics. Please ensure that these are already installed.

For the first step of data download we will use the `galah` package that allows downloading data from several repositories of biodiversity records, including the Atlas of Living Australia. 

For the second step of data preparation we will use useful data wrangling packages `dplyr` and `lubridate`. We will use the package `sf`  to transform the records into an spatial object, and `mapview` to inspect them.  

Finally, we will use `redlistr` to calculate the spatial metrics for the assessment. We also use functions from the `units` package to calculate areas in different units.

```{r Loading packages, message=FALSE}
library(galah)
library(sf)
library(mapview)
library(dplyr)
library(lubridate)
library(units)
library(redlistr)
```

## Step 1: Download the data 

We configure the `galah` package to query [ALA](www.ala.org.au) and provide basic login information. You will need to replace `<YOUR USER NAME>` and `<YOUR EMAIL>` with the values you would use in the ALA portal. 

```{r config galah, echo=FALSE, eval=TRUE}
galah_config(atlas = "Australia",
             username = "neomapas",
             email = "jr.ferrer.paris@gmail.com") 
```

```{r config galah placeholder, echo=TRUE, eval=FALSE}
galah_config(atlas = "Australia",
             username = "<YOUR USER NAME>",
             email = "<YOUR EMAIL>")
```

We make a first query to look at the number of records per data resource. We construct this query using a sequence of function calls concatenated by a "pipe" (`|>`):

```{r how many records, message=FALSE}
galah_call() |>
        identify("Paralucia spinifera") |>
        group_by(dataResourceName) |>
        atlas_counts()
```

We want to replicate the methods used in the most recent assessment. We construct a query to download records from the *NSW BioNet Atlas* and *iNaturalist Australia* and to only include records prior to 2024:

```{r download filtered records, message=FALSE}
selected <- c("NSW BioNet Atlas", "iNaturalist Australia ")
ala_records <- galah_call() |>
        identify("Paralucia spinifera") |>
        filter(
            dataResourceName %in% selected,
            year < 2024) |>
        atlas_occurrences()
```

We want to check that were working with valid presence records:

```{r check records}
ala_records |> 
    group_by(basisOfRecord, occurrenceStatus) |> 
    summarise(n(), .groups="drop")
```

That looks fine, we are ready for the next step.

**Note**: Keep in mind that many biodiversity repositories have rules for sharing coordinates of threatened species. In those cases the coordinates might be obscured. If that is the case for the species of interest, you will need to follow an extra procedure to get access to the raw coordinates of the records.

## Step 2: Prepare the data

The downloaded dataset has few columns with the basic record information:

```{r glimpse at the data}
glimpse(ala_records)
```

We will be working with dates and geospatial coordinates, so we need to do extra steps to transform these columns to special data formats. 

First, we use the function `st_as_sf` to convert the table of records into a spatial points dataset. We specify the columns with the coordinates and the coordinate reference system information corresponding to unprojected geographic coordinates.

Second, we use the `mutate` and `year` functions to extract the year from the date string

```{r prepare dataset}
ala_points <- st_as_sf(ala_records,
    coords = c("decimalLongitude", "decimalLatitude"),
    crs = "EPSG:4326") |>
    mutate(Year=year(eventDate))

```

We want to verify the accuracy of the records we are using, and exclude any dubious records. Erroneous records can arise from misidentifications in the field, mislabeling in the original source or the collection, or by errors introducing during and after digitalisation of the record. 

In real life cases, there should be a comprehensive verification from the source of the raw data to the digital record to flag and correct errors. For this excercise we will restrict ourselves to the filtering of spatial outliers that we classify as dubious records.

We can visualise the records and check for obvious outliers. With the interactive interface of `mapview` we can click on suspicious points and get the `recordIDs`:

```{r visualise records}
mapview(ala_points, zcol = "Year")
```

I suspect that these two records are dubious: 

```{r dubious records}
dubious_records <-  c(
    "7a782f5c-55a5-4863-9644-15e2e3ed79c5",
    "a8f7ab6f-45aa-47bf-a522-a28853ef25e0"
)
```

We can apply the filter function to the spatial dataset in order to get a cleaned list of record:

```{r cleaned dataset}
cleaned_points <- ala_points |>
    filter(!recordID %in% dubious_records)
```


Now we have cleaned the dataset but for the next step we still need to do one more task. The spatial functions of the `redlistr` package require the dataset to be projected in an appropriate coordinate reference system that allow area calculations. 

We can use the function `st_transform` to reproject the data:

```{r projected dataset}
projected_points <- 
    st_transform(cleaned_points,
    "EPSG:3112")

```

Now we can do the next step.

## Step 3: Calculate the metrics

Our data is in a format that can be directly used by the functions in the `redlistr` package. 

The `getEOO` function will create a convex hull around the presence records:

```{r EOO all records}
best_EOO <- getEOO(projected_points)
```

And we can summarise the results using the `summary` function:

```{r summary EOO}
summary(best_EOO)
```

We might want to recalculate EOO using a subset of records. For example, this species shows fluctuations in population size and might undergo local extinctions. Older records without recent resightings might inflate the EOO. 

Here we filter the data, recalculate the EOO and call the summary function in one chunk of code:

```{r alternative EOO}
projected_points |>
    filter(Year>2010) |>
    getEOO() |>
    summary()
```

This example shows that the value of EOO is sensitive to the assumption that all historical records represent the current extent of the species. By excluding older occurrence records we are trying to account for potential local extinctions in the estimate of EOO.

Next we calculate the Area of Occupancy with the function `getAOO`. For species assessments we use a grid size of 2000 m and we don't want to use the 1% rule. 

```{r default AOO}
simple_AOO <- getAOO(projected_points, 
    grid.size=2000,
    bottom.1pct.rule = FALSE,
    jitter = FALSE)
```

The `summary` function provides the results in number of grid cells:
```{r summary AOO}
summary(simple_AOO)
```

The value of the AOO is known to change depending on the placing of the grid. We can use an option called `jitter` to recalculate the AOO multiple times and summarise the range of potential values of AOO.
```{r find best AOO}
best_AOO <- getAOO(projected_points, 
    grid.size=2000,
    bottom.1pct.rule = FALSE,
    jitter = TRUE,
    n_jitter = 99
    )
```

We can use this code to calculate the result in area of occupancy by multiplying the number of cells with their area in square km. 
```{r range of values AOO}
gridsize <- set_units(best_AOO@params$gridsize,'m')^2 |> set_units('km2')
AOOvals <- best_AOO@AOOvals * gridsize
summary(AOOvals)

```

With the use of the jitter option, we get values between 148 and 188 $\mathrm{km}^2$.  

Similarly to the EOO, we might want to recalculate the AOO using subsets of the data. But we will leave this as an excercise for the reader.

Finally, we can now visualise the results together with the input data to verify that they are all consistent. We use again the interactive functions in the `mapview` package:

```{r visualise results}
mapview(projected_points, hide=TRUE) +
mapview(best_AOO@grid) +
mapview(best_EOO@pol)
```

## Step 4: Apply thresholds

Criterion B of the IUCN Red List of Threatened Species is based on the geographic range in the form of either extent of occurrence (B1) or area of occupancy (B2) combined with some conditions that make the species prone to abrupt changes.

These criteria have different quantitative thresholds for each category:

<p style="text-align:center;">

|  Category | Extent of occurrence (EOO) | Area of occupancy (AOO) | Other Conditions |
| ------------- | -------------  | -------------  | -------------  |
| Vulnerable | <20,000 $\mathrm{km}^2$ | < 2,000 $\mathrm{km}^2$ | at least 2 of: a) fragmented or <= 10 locations, B) continuing decline, or c) extreme fluctuations |
| Endangered | <5,000 $\mathrm{km}^2$ | < 500 $\mathrm{km}^2$ | at least 2 of: a) fragmented or <= 5 locations, B) continuing decline, or c) extreme fluctuations |
| Critically Endangered | <100 $\mathrm{km}^2$ | < 10 $\mathrm{km}^2$ | at least 2 of: a) fragmented or one locations, B) continuing decline, or c) extreme fluctuations |
</p>

Our calculation found that the EOO is above 100 $\mathrm{km}^2$ but below 5,000 $\mathrm{km}^2$. From the previous assessment we also know that the species meets the condition of a severely fragmented distribution, and there is evidence of inferred declines in area, extent and/or quality of habitat, and extreme fluctuations in the number of mature individuals. The outcome would be a category of Endangered.

Similarly, our calculation resulted in an AOO above 10 $\mathrm{km}^2$ but below 500 $\mathrm{km}^2$. From the previous assessment we also know that the species meets the condition of a severely fragmented distribution, and there is evidence of inferred declines in area, extent and/or quality of habitat, and extreme fluctuations in the number of mature individuals. The outcome would be a category of Endangered.

## Conclusions

We will compare the outputs of this assessment with the existing assessment for this species:

The values of AOO reported in the conservation assessment of the species was 152 — 176 $\mathrm{km}^2$ and the extent of occurrence (EOO) is 1,693 — 1,823 $\mathrm{km}^2$. The AOO was based on 2 x 2 km grid cells, and the EOO is based on a minimum convex polygon enclosing all mapped occurrences of the species. AOO and EOO and were calculated using ArcGIS software and based on cleaned spatial datasets from the same dataset considered here. AOO and EOO were provided as a range to accommodate the uncertainty around several sites at which local extinction has been documented. 

Differences between the published assessment and our calculations are expected. We tried to use the same data sources and clean the data following similar principles, but we did not use the exact same dataset, and we did not consult experts in the steps of data preparation, so we might be using a different subset of the available data.

Our results estimated a wider range of AOO values (`r min(AOOvals)` — `r max(AOOvals)` $\mathrm{km}^2$) but slightly lower EOO values (1,414 — 1,566 $\mathrm{km}^2$). These differences are well within the thresholds of the category, so they wouldn't affect the outcome of the assessment. 