---
title: "Working with raster and vector data"
subtitle: "Ecosystem distribution data in raster and vector format"
author: "Aniko Toth, José R. Ferrer-Paris"
date: "2026-02-17"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with raster and vector data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

## 1. Introduction

In this vignette, we provide an example of using the tools within
`redlistr` to assess the spatial criteria of the [IUCN Red List of
Ecosystems](https://iucnrle.org/) using different spatial data format. 

As these criteria assess the change in the extent of an ecosystem over 
time (Criterion A) and properties of the geographic distribution size 
(Criterion B). Both of these criteria require the use of maps of ecosystem 
distributions. Given the great variety of spatial data format we can 
expect that different users will be presented with different choices
to describe the historical, current or future distribution of ecosystem 
types.

Here we will explain steps for reading raster and vector maps using functions 
in packages `terra` and `sf`. Please ensure that these are already installed.

```{r Loading packages, message=FALSE}
library(sf)
library(terra)
library(redlistr)
```

## 2. Working with data in raster format

### Importing data 

To import your own raster data, you can use the `terra::rast` function,
which can handle many different types of georeferenced spatial data.
More information on the `terra` package can be found
[here](https://rspatial.github.io/terra/).

Our main example is based on mangrove distributions from the northern 
regions of Western Port Bay, Victoria, Australia. The two raster files
for this example are included with the package, the distribution for 
the year 2000 is based on Giri et al., ([2011](https://onlinelibrary.wiley.com/doi/10.1111/j.1466-8238.2010.00584.x/abstract)). 
The second distribution is
a classification map generated using the `XGBoost` package and data from
Landsat 8, and represents mangrove distribution in 2017. 

Both of these geographic distributions are stored as rasters in GeoTIFF format, 
and a raster value of 1 denotes presence. GeoTIFF is one of the most common
formats for raster data, but functions in `terra` package should be able to 
handle many other formats.

```{r Loading our example distributions}
mangrove.2000 <- rast(system.file("extdata", "example_distribution_2000.tif", 
                                    package = "redlistr"))
mangrove.2017 <- rast(system.file("extdata", "example_distribution_2017.tif", 
                                    package = "redlistr"))
```

### Getting basic geospatial information 

One of the first steps in data preparation is to check basic geospatial
information of the input files. Particularly we want to check resolution 
(grain size), extent, and spatial properties.

Discrepancies in these properties could invalidate spatial analysis and 
will often raise warnings or errors. So it is important to check this first
and take measures to fix any problems before proceeding with other steps.

For spatial analysis, the coordinate reference system (CRS) of the datasets must
be measured in metres, and must be consistent between layers you wish to
compare.

```{r Checking CRS}
crs(mangrove.2000) 
```

Reprojection can be done by using the `terra::project` function. Be sure
to choose a coordinate reference system that is appropriate to your
study area.

### Plotting out data

There are many options to plot and visualise spatial data. The most simple
`plot` function will give us a first preview, and it allows superimposing
multiple objects in one single plot:

```{r Plotting the two rasters, fig.show='hold', fig.width=7, fig.height=7, fig.alt='Overlay of two mangrove distribution maps'}
plot(mangrove.2000, col = "grey30", legend = FALSE, main = "Mangrove Distribution")
plot(mangrove.2017, add = T, col = "darkorange", legend = FALSE)
```

At this stage it is important to check that the data are georectified
(in the right location on Earth). This can be achieved by simply
plotting the map and checking the coordinates. Otherwise, check the
distribution maps against satellite images, perhaps by using packages
such as `ggmap`, `plotgooglemaps`, `googleVis` etc. It is also important
to check that the data are suitable for the task at hand.

### Calculating spatial metrics 

We start by calculating the area of the ecosystem. Area is calculated
here using the pixel count method, where the number of pixels classified
as the ecosystem of interest is multiplied by the area of each pixel,
giving the area of the ecosystem distribution. The area calculated is
provided in square kilometres.

```{r Calculate area of rasters}
a.2000 <- getArea(mangrove.2000)
a.2000
a.2017 <- getArea(mangrove.2017)
a.2017

# you can also calculate areas for both layers at once by combining them into a raster stack:

a <- getArea(c(mangrove.2000, mangrove.2017))
```

The results: at time 1 (2000), the ecosystem was `r a.2000$area` km2,
and at time 2, (2017) the ecosystem was `r a.2017$area` km2.

### Rasters with multiple ecosystems or layers

The `getArea` function is capable of processing rasters with multiple
ecosystems and layers. Layers usually represent time slices. The
function will return a table of areas, with a row for each ecosystem
represented in the raster, and a separate row for each layer if there
are multiple layers.

If you prefer to focus on a single ecosystem or raster value, and you do
not know the value which represents your target ecosystem, you can plot
the ecosystem and use the `terra::click` function. Once you know the
value which represents your target ecosystem, you can create a new
raster object including only that value, or for smaller rasters, simply
filter your output table to the value(s) of interest. Here's an example
with a dummy raster.

```{r rasters with multiple values, eval=TRUE}
# Load dummy multilayer raster for example
rstack <- rast(system.file("extdata", "example_raster_stack.tif", 
                                    package = "redlistr"))

# If the target ecosystem is represented by value = 5
r.bin <- rstack == 5 # Has values of 1 and 0
# Convert 0s to NAs, if desired. 
values(r.bin)[values(r.bin) != 1] <- NA
# get area
getArea(r.bin)

# Alternatively: 
getArea(rstack) |> 
  dplyr::filter(value == 5)
 # note this alternative is slower for large rasters. 
```

The `getAreaTrend` function takes one multilayer raster. It calculates the area for each layer or list element, for each ecosystem. The function returns a list of `trend`
class objects, with one element for each ecosystem or raster value in
the input data.

```{r temporal trends with raster stacks, eval=TRUE}
#input multilayer raster 
rstack <- rast(system.file("extdata", "example_raster_stack.tif", 
                                    package = "redlistr"))
# returns a list of trends, one per ecosystems
area.trend <- getAreaTrend(rstack)

```

Now we can summarise the change for ecosystem 1, this function also produces a 
plot of area over time:

```{r summary of trends, eval=TRUE, fig.alt='Plot of area over time for one ecosystem'}
summary(area.trend[[1]])
```

And we can plot the difference between the two ecosystems over the two time points.

```{r plot of differences, eval=TRUE, fig.alt='Plot of differences in ecosystem distributions between two times'}
plot(area.trend[[1]]@diff)

```

## 3. Converting from raster to vector format

Functions in this package, like the `getEOO` and `getAOO` functions, 
work on polygons as well. You can use functions in packages `terra` and `sf`
 to convert your data from raster to vector format. For vector format you can 
  specify which column contains ecosystem names. If all polygons
are from one ecosystem, you can leave out this argument.

```{r EOO and AOO with polygons, eval = TRUE}
mangrove.2000.poly <- as.polygons(mangrove.2000) |> st_as_sf() 
#fix crs encoding
st_crs(mangrove.2000.poly)$wkt <- enc2utf8(st_crs(mangrove.2000.poly)$wkt)

EOO.poly <- getEOO(mangrove.2000.poly)
AOO.poly <- getAOO(mangrove.2000.poly)
```

## 4. Working with data in vector format

### Importing data as shapefile or .KML file

To import data from a shapefile (.shp) or other common vector formats, 
we can use the `sf::st_read` function.

```{r Importing shapefile, eval=FALSE}
my.shapefile <- st_read('./path/to/folder/shapefile.shp')
my.KML.file <- st_read('./path/to/folder/kmlfile.kml')
```

Once these files are imported, they can then be converted into raster
format via the `terra::rast` or `terra::rasterize` functions if desired.
However, many of the functions in the package will accept vector format
directly.

### Polygon data with multiple ecosystems

If your input data is in polygon format, you can enter the name of the
column containing ecosystem labels into the names_from argument. If left
empty, for polygon inputs the function will treat all polygons as part
of a single ecosystem type.

```{r polygon data for multiple ecosystems, eval = TRUE}
# polygon example
polygons <- st_read(system.file('extdata/', 'example_polygons.shp',
  package = "redlistr"))
poly_areas <- getArea(polygons, names_from = "time1")
```

The `getAreaChange` function takes two inputs in the same format. It
calculates the difference in area between two datasets, which can
contain multiple ecosystems. It can accept data from `data.frame` such
as those output by the `getArea()` function, `SpatRaster`, `SpatVector`
or `sf` objects with POLYGON geometries. Make sure to provide area
measure in km2 if you are using a data.frame as the input. Note that
when providing data frame and polygon inputs, the column name with
unique ecosystem labels must be included in the function call. If the
column name is different in the two inputs, both names must be provided
in the same order as the input datasets.

```{r Using getAreaChange, eval = TRUE}
# input rasters
area.change <- getAreaChange(mangrove.2000, mangrove.2017)

# input data frames 
# value is the column that identifies the ecosystem in both tables.

area.change <- getAreaChange(a.2000, a.2017, value) 

# input polygons 
polygonsT5 <- st_read(system.file('extdata/', 'example_polygons_t5.shp',
  package = "redlistr"))
area.change <- getAreaChange(polygons, polygonsT5, 
  names_from_x = "time1", 
  names_from_y = "time5")

```


## 5. Final words

This vignette includes a range of examples of how to use different input formats with the functions of this package. Please visit the rest of the documentation to learn more about the functions used here, and to follow complete examples of ecosystems and species assessments. 